### 2024-08-12 docs/SOSP 2024/intro.tex
Fix typos

This paper presents \Treebeard{}, a novel schedule-guided compiler
for decision tree inference on multiple hardware targets. \Treebeard{} is able to 
generate high-performance code for decision tree inference by exploring a large optimization space. 
%% RG - rewording
%% The code generation is guided by a \emph{schedule} written in a custom scheduling language that can represent a wide range of implementation strategies. 
This is achieved by a compilation framework consisting of a 
custom scheduling language that can represent a wide range of 
implementation strategies and techniques to efficiently explore
the optimization space. We demonstrate that the language is 
is capable of expressing all optimizations proposed by prior work and more.
Further, our schedule exploration heuristic can quickly find a near 
optimal schedule for the model being compiled. 
%% Further we generalizes them and incorporates several new optimizations. 
% \TODO{RG: it would be good to expand on this and also talk about the retargetable component. 
% We could also say that the schedule framework and the retargetable compiler work in an 
% interwined manner, each benefitting from the other.}  
The second component of \Treebeard{} is a \emph{retargetable} multi-level compiler that can generate efficient code 
for any specified schedule for both CPUs and GPUs. For this purpose, we re-architect \TreebeardOLD{} to
support schedule-guided code generation, and incorporate several new optimizations that are critical to target GPUs. 
These two components of \Treebeard{} are intertwined, each benefitting from the other.
%Last, to address the challenge of selecting a near-optimal schedule from the vast optimization space, 
%%
%we design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.

@@

%% RG: This may not be needed -- can be removed to save space!
%% \TODO{Oganization}
In summary, this paper makes the following contributions.
\begin{itemize}
  \item We identify that there are many different ways to implement  
  decision tree inference and that the best implementation depends both on the model
  and the target processor. We design a scheduling language that 
  allows us to represent this solution space. 
  \item We propose a retargetable compiler that can generate code as specified by 
  the schedule. The compiler also includes new abstractions for caching,
  reduction and representing models in memory. These are critical to generate 
  efficient GPU code.
  \item Since the scheduling language can represent an unbounded number of schedules, 
  we design and implement techniques to search over this space and find high-performance 
  schedules in tens of seconds. 
  \item We exhaustively evaluate \Treebeard{} and report that it achieves 
  portable performance across a range of hardware targets and consistently outperforms
  state-of-the-art decision tree inference frameworks for both CPU and GPU.
\end{itemize}

>>>>

This paper presents \Treebeard{}, a novel schedule-guided compiler
for decision tree inference on multiple hardware targets. \Treebeard{} is able to 
generate high-performance code for decision tree inference by exploring a large optimization space. 
%% RG - rewording
%% The code generation is guided by a \emph{schedule} written in a custom scheduling language that can represent a wide range of implementation strategies. 
This is achieved by a compilation framework consisting of a 
custom scheduling language that can represent a wide range of 
implementation strategies and techniques to efficiently explore
the optimization space. We demonstrate that the language is 
capable of expressing all optimizations proposed by prior work and more.
Further, our schedule exploration heuristic can quickly find a near 
optimal schedule for the model being compiled. 
%% Further we generalizes them and incorporates several new optimizations. 
% \TODO{RG: it would be good to expand on this and also talk about the retargetable component. 
% We could also say that the schedule framework and the retargetable compiler work in an 
% interwined manner, each benefitting from the other.}  
The second component of \Treebeard{} is a \emph{retargetable} multi-level compiler that can generate efficient code 
for any specified schedule for both CPUs and GPUs. For this purpose, we re-architect \TreebeardOLD{} to
support schedule-guided code generation, and incorporate several new optimizations that are critical to target GPUs. 
These two components of \Treebeard{} are intertwined, each benefitting from the other.
%Last, to address the challenge of selecting a near-optimal schedule from the vast optimization space, 
%%
%we design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.

@@

%% RG: This may not be needed -- can be removed to save space!
%% \TODO{Oganization}
In summary, this paper makes the following important contributions.
\begin{itemize}
  \item We identify that there are many different ways to implement  
  decision tree inference and that the best implementation depends both on the model
  and the target processor. We design a scheduling language that 
  allows us to represent this solution space. 
  \item We propose a retargetable compiler that can generate code as specified by 
  the schedule. The compiler also includes new abstractions for caching,
  reduction and representing models in memory. These are critical to generate 
  efficient GPU code.
  \item Since the scheduling language can represent an unbounded number of schedules, 
  we design and implement techniques to search over this space and find high-performance 
  schedules in tens of seconds. 
  \item We exhaustively evaluate \Treebeard{} and report that it achieves 
  portable performance across a range of hardware targets and consistently outperforms
  state-of-the-art decision tree inference frameworks for both CPU and GPU.
\end{itemize}

### 2024-08-12 docs/SOSP 2024/overview.tex

\section{Compiler Overview}
\label{Sec:Overview}
\Treebeard{} takes a serialized decision tree ensemble as input (For example
XGBoost JSON, ONNX etc.) and generates an optimized inference function. 
\Treebeard{} automatically generates an optimized inference function from 
the serialized model and can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor.

@@

Table \ref{Tab:IRSpecification} lists the operations in the three IRs. 
In HIR, the model is represented as a collection of binary trees. This abstraction
allows the implementation of optimizations that require the manipulation of the model
or its constituent trees. Figure \ref{Fig:HIRExample} shows this representation for a model 
with four trees and how the model can be transformed. We use this model as a running example 
for the rest of the section.
After these model-level optimizations are performed on the HIR, the 
code is lowered to the mid-level IR (MIR) as dictated by a user 
specified \emph{\textbf{schedule}}. The schedule determines how
the iteration space that goes over the trees and input rows is to be traversed
by specifying how loops are to be tiled, which loops are to be
parallelized, which loops are to be mapped to GPU grid and block dimensions etc.
(Details in Section \ref{sec:schedule}). While MIR is a loop-based IR that 
explicitly encodes details of the iteration space has to be traversed, 
it still abstracts details about the in-memory representation of the model.

In the lowered MIR, the compiler uses the \op{reduce} op from 
the \op{reduction} dialect we design (details in Section \ref{sec:reduction})
to represent reduction operations. The lowering of the \op{reduce} operation 
involves introducing temporary buffers and splitting the reduction 
to correctly implement reduction in the presence of 
parallel loops. This process, that we call \emph{legalization}, is 
described in Section \ref{sec:reduction}.

The MIR is then further lowered to a low-level IR (LIR). This is the
level at which the compilation pipeline diverges for different targets. In 
the GPU compilation pipeline, the required memory transfers and kernel
invocations are inserted into the LIR. Additionally, buffers 
to hold model values are inserted and tree operations are lowered to
explicitly refer to these buffers. This lowering is controlled by 
a plugin mechanism where different in-memory representations can 
be added to the compiler by implementing an interface 
(Section \ref{sec:representations}). Vectorization of tree traversals
is also explicitly represented in LIR.

In the remainder of this section, we show using our running example
from Figure \ref{Fig:HIRExample} how different schedules can be used to 
lower the model to MIR and LIR. We also show how the \op{reduce} operation
is lowered and legalized.

First, we consider the the schedule that processes one tree at a time 
for all input rows and unrolls the tree walks for all trees in the model. 
We assume that the trees have been reordered so that all trees with 
the same depth are together as shown in Figure \ref{Fig:HIRExample}.
The schedule splits the loop over the trees into two loops -- one that
iterates over the first two trees and the second that iterates over the
last two trees. The schedule the unrolls the tree walks for each tree.
\begin{lstlisting}[style=c++]
  reorder(tree, batch)  
  split(tree, t0, t1, 2)
  unrollWalk(t0, 1)
  unrollWalk(t1, 2)
\end{lstlisting}

\Treebeard{} represents the set of loops as a tree of loops, where the children 
of a loop node are the nodes for the immediately contained loops. Each schedule 
primitive modifies this tree in some way. Additionally,
the lineage of each of the loops is maintained. This allows 
the compiler to automatically infer the ranges for all loops. To lower the
HIR to MIR, the compiler traverses the schedule tree and generates the
corresponding loops. The MIR for the example model, generated using the above   
schedule is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t0 = 0 to 2 step 1 {
    T = getTree(ensemble, t0)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t1 = 2 to 4 step 1 {
    T = getTree(ensemble, t0)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
Subsequent lowering rewrites the \op{walkDecisionTree} operations to a 
series of \op{traverseTile} operations, in this case one for the 
first instance and two for the second (as specified by the unroll depth
attribute). \Treebeard{} also determines that the \op{reduce} operation
can be implemented by a simple addition operation as there are no 
surrounding parallel loops (The legalization of reductions is described 
in detail in Section \ref{sec:reduction}).

We now show how we would compile our example model to a GPU. 
Additionally, we divide the trees into two sets and process each set
in parallel. We also unroll the tree walks for each tree.
The schedule to accomplish this is as follows.
\begin{lstlisting}[style=c++]
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  split(t0, t0_1, t0_2, 2)
  unrollWalk(t0_1, 1)
  unrollWalk(t0_2, 2)
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule processes on input row per thread block 
and splits the trees into two sets, each of which contains 
two trees. The MIR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

In the case of this example, the lowering passes for the \op{reduce}
operation determines that several paralel iterations of the \op{t1} 
loop accumulate into the same element of the \op{result} array.
Therefore, the reduction is rewritten so that each parallel 
iteration accumulates into a different array element by introducing 
a temporary buffer as follows.
\begin{lstlisting}[style=c++]
  float tempResults[2][BATCH_SIZE]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_1 + t1)
        treePred = walkDecisionTree(T, 
                      input[batch]) <unrollDepth = 1>
        reduce(tempResults[t1][batch], treePred)
      }
      for t0_2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_2 + t1)
        treePred = walkDecisionTree(T,
                      input[batch]) <unrollDepth = 2>
        reduce(tempResults[t1][batch], treePred) <'+', 0.0>
      }
    }
    result[batch] = reduce_dimension(result[0], 1)
  }
\end{lstlisting}
Here, partial results are accumulated into \op{tempResults} and then
reduced across the \op{t1} dimension (represented by the \op{reduce\_dimension}
where the second argument represents the dimension number to reduce across)
to get the final result.

Finally, since this computation is being targeted to the GPU, the 
parallel loops are rewritten into kernel calls and the  required 
GPU memory allocations and memory transfers are introduced.

As is evident from these examples, the structure of the loop 
nest in the inference routine can get quite complex even 
for simple optimizations. Writing these routines by hand 
is error prone and time-consuming. Building a principled code 
generator for these routines is the best way to explore the
vast design space. 
% Show simple one tree lowering for CPU with walk unrolling
%    Schedule, MIR, 
% Show parallelization across trees and GPU target

To reiterate, the following are the salient points of \Treebeard{}'s design.
\begin{enumerate}
  \item The compiler uses three intermediate representations (IRs) to
  represent the inference computation at different levels of abstraction. This
  allows different optimizations to be performed and also allows us to share
  infrastructure between compilation pipelines for different target processors.

  \item The specification of how the inference computation is to be lowered to 
  loops is not encoded directly in the compiler. Instead, this 
  is specified as an input to the compiler using a scheduling language that is 
  specialized for decision tree inference computations (Section \ref{Sec:SchedulingLang}).
  This separation allows us to build optimizations and schedule exploration 
  mechanisms independent of the core compiler (Section \ref{AutoTuneHeuristic}). 
  The scheduling language also exposes details like parallelization, optimization
  of reductions (vectorization, use atomics, shared memory on GPUs etc.).

  \item \Treebeard{} has been designed to keep the optimization passes and code 
  generator independent of the in-memory representation finally used for the model. To achieve 
  this, \Treebeard{} specifies an interface to implement that provides the necessary 
  capabilities to the code generator as a plugin. This interface abstracts several details
  on how model values are stored. In particular, it abstracts the actual layout of 
  the memory buffers, how to load model values like thresholds and feature indices, how 
  to move from a node to its children and determining whether a node is a leaf.
  This design allows us to write each memory representation as a standalone plugin 
  and reuse the rest of the compiler infrastructure.
\end{enumerate}

>>>>

\section{Compiler Overview}
\label{Sec:Overview}
\Treebeard{} takes a serialized decision tree ensemble as input (E.g.:
XGBoost JSON, ONNX etc.) and automatically generates an optimized inference function. 
\Treebeard{} automatically generates an optimized inference function from 
the serialized model and can either target CPUs or GPUs. 
Figure \ref{Fig:CompilerStructure} shows the structure of the \Treebeard{} compiler. 
The inference computation is lowered through three intermediate representations
-- high-level IR (HIR), mid-level IR (MIR) and low-level IR (LIR). The LIR is
finally lowered to LLVM and then JIT'ed to the specified target processor.

@@

Table \ref{Tab:IRSpecification} lists the operations in the three IRs. 
In HIR, the model is represented as a collection of binary trees. This abstraction
allows the implementation of optimizations that require the manipulation of the model
or its constituent trees. Figure \ref{Fig:HIRExample} shows this representation for a model 
with four trees and how the model can be transformed. The model contains two trees of 
depth 1 and two trees of depth 2 (\circled{A}). The right side of the diagram shows the models after 
padding and reordering (\circled{B}). Padding makes all leaves in a tree the same depth (Trees 2 and
4 are padded). Trees are reordered so that trees of the same depth are together (Trees 
2 and 3 are swapped). We use this model as a running example for the rest of the section.
After these model-level optimizations are performed on the HIR, the 
code is lowered to the mid-level IR (MIR) as dictated by a  
\emph{\textbf{schedule}}. The schedule determines how to traverse
the iteration space that goes over the trees and input rows 
by specifying how loops are to be tiled, 
parallelized, mapped to GPU grid and block dimensions etc.
(Details in Section \ref{sec:schedule}). While MIR is a loop-based IR that 
explicitly encodes details of the iteration space has to be traversed, 
it still abstracts details about the in-memory representation of the model.

In the lowered MIR, the compiler uses the \op{reduce} op from 
the \op{reduction} dialect we design (details in Section \ref{sec:reduction})
to represent reduction operations. The lowering of the \op{reduce} operation 
involves introducing temporary buffers and splitting the reduction 
to correctly implement reduction in the presence of 
parallel loops. This process, that we call \textbf{\emph{legalization}}, is 
described in Section \ref{sec:reduction}.

The MIR is then further lowered to a low-level IR (LIR). This is the
level at which the compilation pipeline diverges for different targets. In 
the GPU compilation pipeline, the required memory transfers and kernel
invocations are inserted into the LIR. Additionally, buffers 
to hold model values are inserted and abstract tree operations are lowered to
explicitly refer to these buffers. This lowering is controlled by 
a plugin mechanism which enables different in-memory representations to 
be added to the compiler by implementing an interface 
(Section \ref{sec:representations}). Vectorization of tree traversals
is also explicitly represented in LIR.

In the remainder of this section, we show using our running example
from Figure \ref{Fig:HIRExample} how different schedules can be used to 
lower the model to MIR. We also show how the \op{reduce} operation
is lowered and legalized.

First, we consider the schedule that processes one tree at a time 
for all input rows and unrolls all tree walks. 
We assume that the trees have been reordered so that all trees with 
the same depth are together as shown in Figure \ref{Fig:HIRExample}.
The schedule splits the loop over the trees into two loops -- one that
iterates over the first two trees (Trees 1 and 3 with depth 1) and 
the second that iterates over the last two trees (Trees 2 and 4 with
depth 2). The schedule then unrolls the tree walks for each tree.
\begin{lstlisting}[style=c++]
  reorder(tree, batch)  
  split(tree, t0, t1, 2)
  unrollWalk(t0, 1)
  unrollWalk(t1, 2)
\end{lstlisting}

\Treebeard{} represents loops as nodes in a tree where the children
of a node represent immediately contained loops. Each schedule 
primitive modifies this tree in some way. For example, the \op{split}
primitive above duplicates the subtree rooted at the node that is 
being split. The compiler tracks the lineage of each of the loops. This allows 
the compiler to automatically infer the ranges for all loops. To lower the
HIR to MIR, the compiler traverses the tree of loops and generates the
appropriate loops. The MIR for the example model, generated using the above   
schedule is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  for t0 = 0 to 2 step 1 {
    T = getTree(ensemble, t0)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T, 
                    input[batch]) <unrollDepth = 1>
      reduce(result[batch], treePred)
    }
  }
  for t1 = 2 to 4 step 1 {
    T = getTree(ensemble, t0)
    for batch = 0 to BATCH_SIZE step 1 {
      treePred = walkDecisionTree(T,
                    input[batch]) <unrollDepth = 2>
      reduce(result[batch], treePred) <'+', 0.0>
    }
  }
\end{lstlisting}
Subsequent lowering rewrites the \op{walkDecisionTree} operations to a 
series of \op{traverseTile} operations -- one for the 
first instance and two for the second (as specified by the unroll depth
attribute). \Treebeard{} also determines that the \op{reduce} operation
can be implemented by a simple addition operation as there are no 
surrounding parallel loops (The legalization of reductions is described 
in detail in Section \ref{sec:reduction}).

We now show how easily the same model can be compiled to target a GPU 
using \Treebeard{}'s scheduling language. 
A possible schedule to accomplish this is as follows.
\begin{lstlisting}[style=c++]
  tile(tree, t0, t1, 2)
  reorder(batch, t1, t0)
  split(t0, t0_1, t0_2, 2)
  unrollWalk(t0_1, 1)
  unrollWalk(t0_2, 2)
  gpuDimension(batch, grid.x)
  gpuDimension(t1, block.x)
\end{lstlisting}
This schedule processes one input row per thread block (since the \op{batch}
loop is mapped directly to \op{grid.x}).
It also splits the trees into two sets by tiling the \op{tree} loop.
Each of the two sets is processed in parallel. We unroll the tree walks 
for each tree. The MIR generated is as follows. 
\begin{lstlisting}[style=c++]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_1 + t1)
        treePred = walkDecisionTree(T, 
                        input[batch]) <unrollDepth = 1>
        reduce(result[batch], treePred)
      }
      for t0_2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_2 + t1)
        treePred = walkDecisionTree(T,
                        input[batch]) <unrollDepth = 2>
        reduce(result[batch], treePred) <'+', 0.0>
      }
    }
  }
\end{lstlisting}

In the case of this schedule, the lowering passes for the \op{reduce}
operation determines that parallel iterations of the \op{t1} 
loop accumulate into the same element of the \op{result} array.
Therefore, the reduction is rewritten so that each parallel 
iteration accumulates into a different array element by introducing 
a temporary buffer (\op{tempResults}) as follows.
\begin{lstlisting}[style=c++]
  float tempResults[2][BATCH_SIZE]
  model = ensemble(...)
  par.for batch = 0 to BATCH_SIZE step 1 <grid.x> {
    par.for t1 = 0 to 2 step 1 <block.x> {
      for t0_1 = 0 to 2 step 2 {
        T = getTree(ensemble, t0_1 + t1)
        treePred = walkDecisionTree(T, 
                      input[batch]) <unrollDepth = 1>
        reduce(tempResults[t1][batch], treePred)
      }
      for t0_2 = 2 to 4 step 2 {
        T = getTree(ensemble, t0_2 + t1)
        treePred = walkDecisionTree(T,
                      input[batch]) <unrollDepth = 2>
        reduce(tempResults[t1][batch], treePred) <'+', 0.0>
      }
    }
    result[batch] = reduce_dimension(tempResults[:][batch], 0)
  }
\end{lstlisting}
Here, partial results are accumulated into \op{tempResults} and then
reduced across the \op{t1} dimension (represented by the \op{reduce\_dimension}
operation) to get the final result. Finally, since this computation is 
being targeted to the GPU, the parallel loops are rewritten into kernel calls 
and the required GPU memory allocations and memory transfers are introduced.

As is evident from these examples, the structure of the loop 
nest in the inference routine can get quite complex even 
for simple schedules. Writing these routines by hand 
is error-prone and time-consuming. Building a principled code 
generator to automatically generate these routines is the 
best way to explore the vast design space. The key features in 
the design of \Treebeard{} that enable us to build such a code
generator are as follows.

\begin{enumerate}
  \item The compiler uses three intermediate representations (IRs) to
  represent the inference computation at different levels of abstraction. This
  allows different optimizations to be performed and also allows us to share
  infrastructure between compilation pipelines for different target processors.

  \item The specification of how the inference computation is to be lowered to 
  loops is not encoded directly in the compiler. Instead, this 
  is specified as an input to the compiler using a scheduling language that is 
  specialized for decision tree inference computations (Section \ref{sec:schedule}).
  This separation allows us to build optimizations and schedule exploration 
  mechanisms independent of the core compiler (Section \ref{sec:exploring}).

  \item \Treebeard{} has been designed to keep the optimization passes and code 
  generator independent of the in-memory representation finally used for the model. To achieve 
  this, \Treebeard{} specifies an interface to implement that provides the necessary 
  capabilities to the code generator as a plugin. This interface abstracts several details
  on how model values are stored (Details in Section \ref{sec:representations}). This 
  design allows us to write each in-memory representation as a standalone plugin 
  and reuse the rest of the compiler infrastructure. \TODO{Should we also say it lets us
  easily search over multiple representations?}
\end{enumerate}

### 2024-08-12 docs/SOSP 2024/reduction.tex

\section{Representing and Optimizing Reductions}
\label{sec:reduction}
\Treebeard{} needs to sum up individual tree predictions to compute 
the prediction of the model while performing inference. However,
generating fused reductions within arbitrary loop nests specified 
using \Treebeard{}'s scheduling language is non-trivial. We found 
that existing reduction support in MLIR is insufficient to code
generate and optimize these reductions. MLIR only supports reductions
of value types and does not provide ways to lower reductions to GPUs. 
To address this gap, we design an MLIR dialect that allows us to
specify accumulating values into an element of a multi-dimensional array
and can be lowered to CPU or GPU.

The main abstraction we introduce is the \op{reduce} op. It models 
atomically accumulating values into an element of a 
multi-dimensional array (represented by an MLIR \op{memref}).
Consider the following \Treebeard{} schedule.
In our example, \op{N\_t} is the number of trees 
and \op{batch\_size} is the batch size. The schedule tiles the 
tree loop and parallelizes the resulting outer loop.
\begin{lstlisting}[style=c++]
  tile(tree, t0, t1, N_t/2);
  reorder(t0, t1, batch);
  parallel(t0);
\end{lstlisting}

@@

The compiler simply generates a \op{reduce} op to perform
the required parallel reduction. 
The semantics of the \op{reduce} op is exactly the semantics of 
an atomic accumulation, i.e. it guarantees that all accumulations 
are correctly performed even in the presence of parallel loops. 
The \op{reduce} op is defined for all associative and commutative
reduction operations with a well-defined initial value. The 
reduction operator and the initial value are attributes applied
on the \op{reduce} op.

@@

\subsection{Lowering Reduction Operations}
% We implement lowering of the operations defined above to both the CPU and GPU.
% Since the lowering pipeline from MIR to LIR are different for CPU and GPU 
% compilation, we implement lowering and optimization of our reduction dialect to CPUs and
% GPUs simply using different MLIR rewrite patterns. 
In this section, we briefly describe 
how we lower the reduction operations to the CPU and GPU.

@@

% \subsubsection{Lowering to GPU}
The same reduction abstractions can be lowered to efficient GPU implementations
and therefore, simplify higher-level code generation. 
% The lowering for 
% the inplace and non-inplace operations are essentially the same, except 
% for the target array and we do not distinguish between them except 
% for finally storing the result. 
%
% The lowering of the \op{reduceDim*} 
\Treebeard{} can lower these ops to either exploit
parallelism across the independent reductions or 
the inherent parallelism in the reduction by performing a divide and conquer 
reduction depending on whether there are enough independent reductions to keep all threads
in a thread block busy.
% then the lowering pass can generate code that performs one (or 
% multiple) reductions in each thread. If, however, there are not 
% enough independent reductions, then the lowering pass generates a tree 
% style reduction where multiple threads cooperate to perform a single reduction
% using inter-thread shuffles.

% Another feature specific to GPU reductions is the use of shared memory. 
Additionally, if the schedule specifies that the reduction needs to be performed 
using shared memory, the privatized buffer is allocated in shared memory. 
% The compiler only allocates as much shared memory 
% as needed to hold values processed by a single thread-block.
%  and 
% index offsets are appropriately rewritten to handle the differences between 
% the indexing of the target memref and the shared memory array.
Our abstractions for reductions allow our lowering passes to be 
agnostic of whether we use shared memory and therefore allow 
us to enable or disable shared memory use independently from the other 
parts of the compiler.

>>>>

\section{Representing and Optimizing Reductions}
\label{sec:reduction}
\Treebeard{} needs to sum up individual tree predictions to compute 
the prediction of the model while performing inference. However,
generating fused reductions within arbitrary loop nests specified 
using \Treebeard{}'s scheduling language is non-trivial. We found 
that existing reduction support in MLIR is insufficient to code
generate and optimize these reductions. MLIR only supports reductions
of value types and does not provide ways to lower reductions to GPUs. 
To address this gap, we design an MLIR dialect that allows us to
specify accumulating values into an element of a multidimensional array
and can be lowered to CPU or GPU.

The main abstraction we introduce is the \op{reduce} op. It models 
atomically accumulating values into an element of a 
multidimensional array (represented by an MLIR \op{memref}).
Consider the following \Treebeard{} schedule.
In our example, \op{N\_t} is the number of trees 
and \op{batch\_size} is the batch size. The schedule tiles the 
tree loop and parallelizes the resulting outer loop.
\begin{lstlisting}[style=c++]
  tile(tree, t0, t1, N_t/2);
  reorder(t0, t1, batch);
  parallel(t0);
\end{lstlisting}

@@

The compiler simply generates a \op{reduce} op to perform
the required parallel reduction. 
The semantics of the \op{reduce} op is exactly the semantics of 
an atomic accumulation, i.e. it guarantees that all accumulations 
are correctly performed even in the presence of parallel loops. 
The \op{reduce} op is defined for all associative and commutative
reduction operations with a well-defined initial value. The 
reduction operator and the initial value are attributes
on the \op{reduce} op.

@@

\subsection{Lowering Reduction Operations}
% We implement lowering of the operations defined above to both the CPU and GPU.
% Since the lowering pipeline from MIR to LIR are different for CPU and GPU 
% compilation, we implement lowering and optimization of our reduction dialect to CPUs and
% GPUs simply using different MLIR rewrite patterns. 
% In this section, we briefly describe 
% how we lower the reduction operations to the CPU and GPU.

@@

% \subsubsection{Lowering to GPU}
The same reduction abstractions can be lowered to efficient GPU implementations
and therefore, simplify higher-level code generation. 
% The lowering for 
% the inplace and non-inplace operations are essentially the same, except 
% for the target array and we do not distinguish between them except 
% for finally storing the result. 
%
% The lowering of the \op{reduceDim*} 
\Treebeard{} can lower these ops to either exploit
parallelism across the independent reductions or 
the inherent parallelism in the reduction by performing a divide and conquer 
reduction depending on whether there are enough independent reductions to keep all threads
in a thread block busy.
% then the lowering pass can generate code that performs one (or 
% multiple) reductions in each thread. If, however, there are not 
% enough independent reductions, then the lowering pass generates a tree 
% style reduction where multiple threads cooperate to perform a single reduction
% using inter-thread shuffles.
% Another feature specific to GPU reductions is the use of shared memory. 
Additionally, if the schedule specifies that the reduction needs to be performed 
using shared memory, the privatized buffer is allocated in shared memory. 
% The compiler only allocates as much shared memory 
% as needed to hold values processed by a single thread-block.
%  and 
% index offsets are appropriately rewritten to handle the differences between 
% the indexing of the target memref and the shared memory array.
Our abstractions for reductions allow our lowering passes to be 
agnostic of whether we use shared memory and therefore allow 
us to enable or disable shared memory use independently from the other 
parts of the compiler.
